---
title: "Introduction to CovDecayAnalyzer"
author:
  - name: Carly Bobak
    affiliation: Dartmouth College
package: CovDecayAnalyzer
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to CovDecayAnalyzer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 7,
    fig.height = 4.5,
    fig.align  = "center"
)
```

# Motivation

When a mixed viral or bacterial population is sequenced while it is
actively replicating, read coverage across the reference genome is
usually not uniform. Regions near an active origin of replication
accumulate additional copies during ongoing replication and therefore
higher coverage. Regions far from an origin, or beyond termination
sites, are relatively underrepresented. The *rate* at which coverage
falls off with distance from an origin, or from any other reference
coordinate of interest, can carry biological information about
replication dynamics, and the *difference* in that rate between
experimental conditions can reveal treatment-induced changes.

`CovDecayAnalyzer` provides a small, focused set of tools for this
class of analysis. Given per-base coverage across a linear reference
genome, the package:

1. Bins and median-normalizes coverage per sample.
2. Fits log-linear decay slopes with distance from a user-supplied
   anchor position (origin of replication, genome end, feature
   boundary, or any other coordinate).
3. Compares decay slopes across treatment groups with a linear
   mixed-effects model, tested with a Kenward-Roger small-sample F
   statistic.
4. Assesses whether an observed group difference is specific to the
   anchor of interest by comparing it to a null distribution of
   F-statistics built from random genomic anchor positions.

# When this package is likely to help

Concrete examples of the kind of question the package is built for:

- **Comparative replication dynamics.** How does coverage decay around
  a known or putative origin differ between drug-treated and untreated
  viruses, wild-type and knockout strains, or immortalized vs.
  primary host cells?
- **Origin discovery, in the coarse sense.** Which of several
  candidate anchor positions produces the strongest between-group
  coverage-decay signal?
- **Genome-end vs. interior effects.** Herpesviruses, phages, and
  linear plasmids sometimes show distinctive coverage patterns near
  their ends. The package's anchor is user-supplied, so genome start
  and end are natural choices as well as OriLyt.
- **General coverage-anomaly comparisons.** Any experiment where the
  question is "does the spatial pattern of coverage around a specific
  reference coordinate differ between two or more experimental
  conditions?"

The package was developed for Human Cytomegalovirus (HCMV) TB40/E work
around the lytic origin of replication (OriLyt), but nothing in the
methods is HCMV-specific. Any linear reference genome with a defined
anchor coordinate is a candidate.

# When to use something else

This is a small package doing a specific job. It is not the right
tool if:

- Your question is about coverage *at* a set of features (peaks,
  binding sites, gene bodies) rather than coverage *decay* from an
  anchor. Look at
  `r BiocStyle::Biocpkg("csaw")`,
  `r BiocStyle::Biocpkg("DESeq2")` on binned counts, or
  `r BiocStyle::Biocpkg("edgeR")` on windows.
- You need per-base statistical modeling. This package works on
  binned coverage.
- Your genome is circular and you need distances to wrap around.
  Distances here are computed as absolute linear distance from the
  anchor. Circular topology is not currently supported.
- You have only one sample per group. The formal test requires random
  slopes per sample, and its small-sample correction is designed for
  the n = 2 to 5 per group regime.

# Installation and setup

The package is not yet on Bioconductor. From GitHub:

```r
# install.packages("remotes")
remotes::install_github("BoscoResearchGroup/CovDecayAnalyzer")
```

```{r load-package, message = FALSE}
library(CovDecayAnalyzer)
```

# Input data format

The package reads tab-delimited per-base coverage files with the
following columns:

| Column               | Type    | Description                              |
|:---------------------|:--------|:-----------------------------------------|
| `position`           | integer | 1-based reference coordinate             |
| `unique_top_cov`     | numeric | Uniquely-mapped reads on the top strand  |
| `unique_bot_cov`     | numeric | Uniquely-mapped reads on the bottom strand |
| `redundant_top_cov`  | numeric | Multi-mapped reads on the top strand     |
| `redundant_bot_cov`  | numeric | Multi-mapped reads on the bottom strand  |

This is exactly the schema produced by breseq's mutation
identification stage as `Exported.coverage.tab`. Additional columns
present in some breseq versions (`raw_redundant_top_cov`,
`raw_redundant_bot_cov`, `e_value`) are accepted and ignored.

Total coverage is computed as the sum of the four uniqueness- and
strand-specific columns. If you have coverage from another tool, save
it in this schema before reading and the rest of the pipeline works
unchanged.

# The shipped example data

Four simulated samples are included in `inst/extdata/` for
demonstration and testing. They are entirely synthetic; nothing here
is derived from real data.

The simulated genome is 2000 bp long with an anchor at position 1000.
Samples fall into two groups:

- **WT-GCV** (samples `A1`, `A2`): coverage decays gently from the
  anchor at rate $5 \times 10^{-4}$ per bp
- **WT+GCV** (samples `B1`, `B2`): coverage decays six times faster
  at rate $3 \times 10^{-3}$ per bp

Both groups peak at coverage ~500 near the anchor, so they differ
*only* in decay rate. This is the exact interaction the package is
designed to detect. Running the pipeline on this data should give a
clear, unambiguous signal in every plot and every test.

```{r load-metadata}
extdata_dir <- system.file("extdata", package = "CovDecayAnalyzer")

meta <- read.csv(
    file.path(extdata_dir, "sample_metadata.csv"),
    stringsAsFactors = FALSE
)
meta
```

# Pointing the loader at coverage files

`load_all_samples()` reads its file paths from a `coverage_file`
column in the sample metadata table. This is a deliberate design
choice: the package does not assume any particular on-disk layout,
which makes it work across output conventions from different tools.

For users with breseq output in the standard layout, the helper
`breseq_coverage_paths()` builds the `coverage_file` column in one
line:

```{r build-paths-breseq}
meta$coverage_file <- breseq_coverage_paths(
    sample_names = meta$sample,
    base_dir     = extdata_dir
)
```

For any other tool or layout, populate `coverage_file` directly:

```{r build-paths-generic, eval = FALSE}
meta$coverage_file <- c(
    "/data/experiment1/sampleA_coverage.tab",
    "/data/experiment1/sampleB_coverage.tab",
    "/data/experiment2/sampleC_coverage.tab",
    "/data/experiment2/sampleD_coverage.tab"
)
```

# Loading and normalizing coverage

```{r load-samples}
combined <- load_all_samples(
    sample_metadata = meta,
    bin_size        = 100,      # small bins for the 2 kb toy genome
    genome_length   = 2000
)
combined
```

Each row is one bin from one sample. Coverage has been aggregated to
100 bp windows and median-normalized within each sample; the
`log_norm_cov` column is the log-transformed normalized coverage used
by downstream modeling.

Bin size is a user choice. Smaller bins retain more spatial detail
but produce noisier per-bin means and larger working tables. Larger
bins are smoother but may average across distinct local features. For
a full HCMV genome (~238 kb), 1000 bp bins are typical; for the small
example genome here, 100 bp bins give enough resolution to see the
decay.

# Visualizing decay

`plot_coverage_decay()` shows per-sample normalized coverage against
distance from the anchor, with a loess-smoothed trend line per group.

```{r plot-coverage-decay}
plot_coverage_decay(
    combined,
    anchor_position = 1000,
    window          = 1000
)
```

The two group curves separate cleanly, matching what was built into
the simulated data. On real data, the loess smoother is useful for
spotting local departures from log-linearity that would violate the
assumptions of the formal test in Section \@ref(formal-group-comparison).

# Per-sample decay slopes as a diagnostic

`fit_decay_slopes()` fits a simple linear regression per sample of
`log_norm_cov ~ distance_from_anchor` and returns per-sample slope
estimates:

```{r fit-slopes}
slopes <- fit_decay_slopes(combined, anchor_position = 1000)
slopes
```

WT+GCV samples show slopes several times more negative than WT-GCV.

```{r plot-slopes}
plot_decay_slopes(slopes)
```

This view is a diagnostic, not a formal test. Per-sample slopes throw
away information about how *precisely* each slope is estimated, which
is what the formal LMM in the next section preserves. Look at this
plot to spot outlier samples and to confirm the direction and rough
magnitude of the effect; use the LMM for inference.

# Formal group comparison

`test_decay_slopes()` fits a linear mixed-effects model to all bins
across all samples simultaneously, and tests whether the rate of
decay differs between groups. The full model, the small-sample
correction, and interpretation guidance are in the companion
vignette; see
`vignette("null-distribution", package = "CovDecayAnalyzer")`.

```{r test-slopes}
res <- test_decay_slopes(combined, anchor_position = 1000)
res
```

The returned `decay_test` object has:

- `res$F_stat`, `res$df1`, `res$df2`, `res$p_value`: the
  Kenward-Roger F test for the interaction.
- `res$group_slopes`: per-group slope estimates with 95% confidence
  intervals.
- `res$pairwise`: Tukey-adjusted pairwise group contrasts on slopes.
- `res$model`: the fitted `lmerModLmerTest` object for further
  diagnostics.

```{r pairwise}
res$pairwise
```

# Anchor-specificity via random-anchor null

A significant group difference at one anchor could arise anywhere in
the genome by chance. `test_decay_significance()` refits the LMM at
`n_random` random anchor positions to build a null distribution of
interaction F-statistics, and reports how surprising the observed F
is relative to that null.

For a real analysis, use `n_random = 500` or more and pass a
`BiocParallel` param object to parallelize. Here we use a smaller
value to keep the vignette build quick:

```{r test-significance, message = FALSE, warning = FALSE}
sig <- test_decay_significance(
    combined,
    anchor_position = 1000,
    genome_length   = 2000,
    n_random        = 50,
    seed            = 2026
)
sig
```

The empirical p-value hits the floor of $1 / (n_{\text{random}} + 1)$,
indicating the observed F exceeds every F obtained at a random
genomic position. On real data, the interpretation depends on both
the LMM p-value at the anchor and the empirical p-value; see the
methods vignette for a table of the four meaningful combinations.

```{r plot-null}
plot_null_distribution(sig)
```

# Viral reads-per-million summary

Independent of the decay analysis, viral RPM (reads per million host
reads) is a commonly-reported summary of viral yield across samples.
`compute_viral_rpm()` computes RPM vectorised over samples given
viral read counts and host read counts, and `plot_rpm_by_group()`
draws a simple by-group barplot with error bars.

Because the shipped example data does not include host read counts,
we illustrate with a small synthetic tibble:

```{r rpm-example}
set.seed(1)
rpm_df <- data.frame(
    sample    = paste0("s", 1:12),
    group     = rep(c("WT-GCV", "WT+GCV", "FV-GCV", "FV+GCV"), each = 3),
    viral_rpm = c(
        rnorm(3, 3000, 500),
        rnorm(3, 1500, 300),
        rnorm(3, 2500, 400),
        rnorm(3, 1200, 250)
    )
)

plot_rpm_by_group(rpm_df)
```

In a typical breseq-based workflow, viral read counts come from the
`total_reads` field of `output/summary.json`, and host read counts
come from the concordant-pair line in bowtie2's alignment log.
Extracting these is a small shell script and is not part of the R
package.

# Adapting to your own data

Nothing in the workflow is specific to HCMV or breseq. To use the
package with other data:

- **Different genome.** Set `genome_length` appropriately and choose
  an `anchor_position` that makes biological sense for your reference.
- **Different sequencing tool.** Ensure your per-base coverage file
  has the five required columns (`position`, `unique_top_cov`,
  `unique_bot_cov`, `redundant_top_cov`, `redundant_bot_cov`). If your
  tool produces coverage in a different schema, a small preprocessing
  script that renames columns and writes tab-delimited output is
  usually sufficient. Then either point `load_all_samples()` at these
  files directly via the `coverage_file` column, or write a
  layout-specific helper analogous to `breseq_coverage_paths()`.
- **Bin size and window.** Larger genomes benefit from larger bins.
  For a 5 Mb bacterial genome, 5-10 kb bins are reasonable. The
  `window` argument to `test_decay_slopes()` restricts the fit to a
  local region around the anchor; use it when you have a specific
  distance range in mind (say, "the first 50 kb"), and leave it
  `NULL` for a whole-genome fit.
- **Number of groups.** The LMM handles any number of groups, not
  just two. Pairwise contrasts from `emmeans` scale accordingly.

# Where to go next

- `vignette("null-distribution", package = "CovDecayAnalyzer")` for
  the statistical methods behind `test_decay_slopes()` and
  `test_decay_significance()`.
- `help(package = "CovDecayAnalyzer")` for the full function
  reference. Every exported function has a runnable example.
- The package source on GitHub for bug reports and feature requests.

# Session info

```{r session-info}
sessionInfo()
```
