---
title: "Statistical methods: group decay tests and anchor-specificity"
author:
  - name: Carly Bobak
    affiliation: Dartmouth College
package: CovDecayAnalyzer
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Statistical methods: group decay tests and anchor-specificity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 7,
    fig.height = 4.5,
    fig.align  = "center"
)
```

# What this vignette is for

`CovDecayAnalyzer` has two testing functions,
`test_decay_slopes()` and `test_decay_significance()`. Between them
they answer two related but distinct questions about coverage decay
around a genomic anchor. This vignette is a companion to
`vignette("intro", package = "CovDecayAnalyzer")` that documents the
underlying statistical procedures.

It is written to serve three audiences:

- Users who want to understand what the tests are actually doing
  before running them on real data.
- Reviewers of downstream manuscripts who need enough detail to
  evaluate the appropriateness of the methods.
- Students or curious readers who want a worked example of applying
  linear mixed-effects models plus Monte Carlo permutation testing to
  a spatial-genomics problem.

The exposition tries to remain honest about what is done and what is
not, including limitations and reasonable alternatives.

# The two questions

Suppose you have several samples in each of two or more treatment
groups, and per-base sequencing coverage on a shared reference
genome. You have a genomic anchor position in mind (an origin of
replication, a feature boundary, a genome end). The two questions
this package answers:

- **Question A: Do groups differ in coverage-decay rate at this
  anchor?** Answered by `test_decay_slopes()`, which fits a linear
  mixed-effects model with a distance-by-group interaction and tests
  the interaction with a Kenward-Roger F statistic.
- **Question B: Is any group difference we see at this anchor
  specific to the anchor, or would we see something comparable at
  other positions in the genome?** Answered by
  `test_decay_significance()`, which repeats the LMM at many random
  anchor positions to build a null distribution of interaction
  F-statistics and reports an empirical p-value.

Both questions matter. A positive answer to A alone tells us the
groups differ *somewhere* along the coverage-vs-distance curve near
this anchor. A positive answer to B is what lets us claim that
whatever effect exists is meaningfully localized to the anchor. See
Section \@ref(interpretation-the-two-p-values-together) for a
walk-through of the four combinations of A and B outcomes.

# The linear mixed-effects test

## Why not a two-stage approach?

The most naive way to test for a group difference in decay is a
two-stage procedure: fit a slope per sample, then compare slopes
across groups with a t-test, ANOVA, or Kruskal-Wallis test. This is
easy to explain but has real weaknesses:

- **It discards precision information.** A slope estimated cleanly
  from a well-covered sample counts the same as a slope estimated
  noisily from a sparse sample.
- **It has low power at typical experimental sizes.** With n = 2 to
  5 samples per group, parametric tests on n slopes have very little
  power, and permutation tests span only a coarse p-value grid.

Fitting a single linear mixed-effects model to the full per-bin data
avoids both issues. It uses every observation, propagates
within-sample uncertainty into the group comparison, and admits a
principled small-sample correction.

## Model specification

For sample $i$, bin $j$, group $g$, define $y_{ij}$ as the normalized
log-scale coverage `log_norm_cov` at bin $j$ of sample $i$, and
$d_{ij}$ as the absolute distance from the anchor position, in base
pairs. The model fit by `test_decay_slopes()` is

$$
y_{ij} =
    (\beta_0 + b_{0i}) +
    (\beta_1 + b_{1i}) \, d_{ij} +
    \sum_{g=2}^{G} \gamma_g \, \mathbf{1}(\text{group}_i = g) +
    \sum_{g=2}^{G} \delta_g \, d_{ij} \, \mathbf{1}(\text{group}_i = g) +
    \varepsilon_{ij}
$$

with

- fixed intercept $\beta_0$ and fixed slope $\beta_1$ for the
  reference group,
- fixed group offsets $\gamma_g$ and group-by-distance interaction
  terms $\delta_g$,
- per-sample random intercepts $b_{0i}$ and random slopes $b_{1i}$
  drawn jointly from an unstructured bivariate normal,
- residuals $\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$
  assumed independent.

In `lme4` notation this is

```
log_norm_cov ~ distance_from_anchor * group +
               (1 + distance_from_anchor | sample)
```

## The hypothesis being tested

"Groups differ in decay rate" corresponds formally to the joint
hypothesis

$$
H_0 : \delta_2 = \delta_3 = \dots = \delta_G = 0
$$

that is, all interaction terms are zero, versus the general
alternative that at least one is nonzero. This is a Type III F test
on the `distance_from_anchor:group` interaction row of the ANOVA
table.

## Small-sample correction

With n = 2 to 5 samples per group, the asymptotic assumptions
underlying naive Wald or Satterthwaite tests are unreliable. The
Kenward-Roger approximation (Kenward and Roger, 1997) adjusts both
the fixed-effect covariance estimate and the denominator degrees of
freedom of the F statistic. It is designed for exactly this
small-sample regime and is the standard recommendation in the mixed
model literature for experiments with few units per group.

`test_decay_slopes()` invokes `lmerTest::lmer` (Kuznetsova, Brockhoff
and Christensen, 2017) on top of `lme4::lmer` (Bates, Machler, Bolker
and Walker, 2015), then extracts the Kenward-Roger F, denominator
degrees of freedom, and p-value from `anova(model, ddf = "Kenward-Roger")`.

## Post-hoc pairwise contrasts

When the joint interaction test is significant, `emmeans::emtrends()`
(Lenth, 2024) computes per-group slope estimates on the natural
scale, and `pairs()` returns Tukey-adjusted pairwise slope contrasts.
These appear in the `group_slopes` and `pairwise` fields of the
returned object.

## Assumptions

The model above assumes:

- **Log-linearity of decay.** Real coverage decay from an origin of
  replication is not always a clean exponential. Assembly artifacts,
  fork termination sites, and cell-cycle asynchrony can produce
  non-log-linear patterns. The loess overlay in
  `plot_coverage_decay()` is a useful visual check: strong
  systematic departures from monotonic decay indicate the log-linear
  assumption is under strain.
- **Sample independence given group.** Samples within a group are
  assumed to be independent replicates. Repeated measures or
  clustered sample designs would need additional random-effect
  structure.
- **Symmetric decay around the anchor.** Distance is computed as
  absolute distance. If you have reason to expect asymmetric decay
  (e.g. directional replication with an obligate leading and lagging
  strand at your chosen anchor), consider fitting the two arms
  separately.
- **Approximate normality of the residuals.** Coverage is a count
  process, so a Gaussian assumption on `log(norm_cov + epsilon)` is
  an approximation. It is a reasonable one at moderate to high
  coverage. Very low-coverage regions may be better handled by a
  generalized linear mixed model, which is not currently implemented.

## Numerical scaling

Raw distances in base pairs can be very large (tens or hundreds of
kilobases for viral genomes) while per-bp slopes are correspondingly
tiny ($10^{-5}$ to $10^{-3}$). The resulting design matrix condition
number is well beyond double precision, and the Kenward-Roger variance
adjustment fails on such inputs with `system is computationally
singular`.

`test_decay_slopes()` handles this by internally rescaling distance
so its maximum lands in the 1-to-10 range before fitting the LMM.
Slope estimates, standard errors, and confidence intervals in the
returned `group_slopes` and `pairwise` tables are converted back to
per-bp units on the way out. F, degrees of freedom, and p-value are
invariant to this rescaling, so nothing about the test itself is
affected. The scaling factor used is recorded in the
`distance_scale` field of the returned object.

## Worked example

We use the shipped simulated data. The two groups differ in decay
rate by a factor of six.

```{r load-data, message = FALSE}
library(CovDecayAnalyzer)

extdata_dir <- system.file("extdata", package = "CovDecayAnalyzer")
meta <- read.csv(
    file.path(extdata_dir, "sample_metadata.csv"),
    stringsAsFactors = FALSE
)
meta$coverage_file <- breseq_coverage_paths(meta$sample, extdata_dir)

combined <- load_all_samples(
    sample_metadata = meta,
    bin_size        = 100,
    genome_length   = 2000
)
```

```{r run-lmm}
res <- test_decay_slopes(combined, anchor_position = 1000)
res
```

The interaction is highly significant. The per-group slopes are
close to the simulated values, roughly $-5 \times 10^{-4}$ for WT-GCV
and $-3 \times 10^{-3}$ for WT+GCV.

```{r pairwise-contrast}
res$pairwise
```

Basic residual diagnostics:

```{r diagnostics-plot, fig.height = 4}
resid_df <- data.frame(
    fitted    = fitted(res$model),
    residual  = resid(res$model)
)
plot(
    resid_df$fitted, resid_df$residual,
    xlab = "Fitted value", ylab = "Residual",
    main = "Residuals vs fitted",
    pch  = 20, col = "grey40"
)
abline(h = 0, col = "red", lty = 2)
```

Residuals should be roughly homoscedastic and centered on zero. Fan
shapes or systematic curvature suggest misspecification.

# The random-anchor null distribution

## Why permute anchor positions rather than sample labels?

There are two natural things to permute in this setting:

1. **Sample labels** (swap which sample is in which group). Under the
   null of "no group effect", this preserves the marginal
   distribution of coverage at every position. The LMM p-value is
   effectively a parametric proxy for this permutation. This is a
   test of "does treatment matter?".
2. **Anchor positions** (keep the samples and labels fixed, but ask
   the same question at other genomic coordinates). Under the null
   of "the anchor is nothing special", we would expect similar
   between-group differences to appear at random positions. This is
   a test of "is the effect specific to this anchor?".

The two answer different questions and are both useful. The LMM
p-value at the observed anchor covers (1). The empirical p-value
from `test_decay_significance()` covers (2).

Permuting anchor positions rather than sample labels is what makes
sense biologically for the second question. If coverage anomalies
that happen to correlate with treatment are common across the genome
(mapping artifacts, GC content, expression differences that shadow
treatment), a group effect at any single anchor is not automatically
biologically meaningful. Comparing to random positions calibrates
against this background.

## Construction

`null_distribution_F()` samples $N$ positions uniformly from
$[1, \texttt{genome\_length}]$, and at each position refits the LMM
above, extracting the Kenward-Roger F for the distance-by-group
interaction. Fits that fail (rare, and typically near the genome
ends where the window is degenerate) are dropped. The returned
vector has length at most $N$.

An optional `exclude` argument lets you specify a region around the
observed anchor to remove from the random-position pool, to prevent
overlap between the observed window and null windows.
`test_decay_significance()` exposes this via `exclude_radius`.

## Empirical p-value

Given the observed F at the anchor of interest, $F_{\text{obs}}$, and
a null distribution $\{F_1, F_2, \dots, F_n\}$, the empirical
right-tailed p-value is

$$
p_{\text{emp}} = \frac{k + 1}{n + 1},
\qquad
k = \# \{ F_i : F_i \ge F_{\text{obs}} \}.
$$

The `+1` in both numerator and denominator is not smoothing. It is
the correct exact expression for a Monte Carlo test where the null
samples are drawn with replacement, as derived by Phipson and Smyth
(2010). Using the more intuitive $k/n$ estimator systematically
understates the true p-value by roughly $1/n$ and can produce
impossible p-values of exactly zero. The corrected form has a
minimum achievable value of $1/(n+1)$, which is the smallest p-value
that a Monte Carlo test with $n$ samples can honestly report.

## Interpretation: the two p-values together

The LMM p-value at the observed anchor and the empirical p-value
answer different questions, and their combinations are meaningful in
different ways:

|                                | Small empirical p                  | Large empirical p                                            |
|:-------------------------------|:-----------------------------------|:-------------------------------------------------------------|
| **Small LMM p (at anchor)**    | Real effect, and it is specific to this anchor. Best case for a biological claim. | Real effect, but not specific to this anchor. Suggests widespread coverage anomalies that happen to correlate with treatment; interpret with caution. |
| **Large LMM p (at anchor)**    | Cannot occur by construction: if the observed F is unremarkable, it cannot be extreme relative to a null derived from data with the same structure. | No detectable group difference in decay at this anchor.      |

The lower-left cell is precluded because the empirical p-value uses
$F_{\text{obs}}$ as the reference; if that reference is not
extreme, no null can make it look extreme.

The upper-right cell is the most nuanced case. In practice, a small
LMM p combined with a large empirical p suggests one of two
possibilities:

- The chosen anchor is not the "right" position, but similar anchors
  nearby (or elsewhere) show comparable effects. A biological
  narrative attached specifically to *this* anchor is weakened.
- There is a real, widespread biological difference in
  coverage-vs-distance structure between groups, not localized to any
  particular position. Interesting, but a different claim.

## Assumptions of the random-anchor null

The random-anchor approach assumes:

- **Positions in the genome are exchangeable for the purpose of the
  null.** If your genome has structural features that make certain
  positions systematically different (e.g. highly repetitive
  regions, low-complexity tracts), the null may include positions
  that are not biologically comparable to the anchor of interest.
  Consider masking or restricting the sampled positions.
- **The LMM is fit-able at random anchors.** In practice a small
  fraction of random anchors near genome ends produce singular fits.
  These are dropped and the effective $n$ is reported.
- **The number of random anchors is large enough.** The minimum
  achievable p-value is $1/(n_{\text{random}} + 1)$. For screens,
  500 is a reasonable default; for publication-grade significance
  claims, 1000 or more is safer.

## Worked example

Continuing from the LMM example above:

```{r run-null, message = FALSE, warning = FALSE}
sig <- test_decay_significance(
    combined,
    anchor_position = 1000,
    genome_length   = 2000,
    n_random        = 100,
    seed            = 2026
)
sig
```

The empirical p-value is at the floor of $1 / (100 + 1) \approx 0.0099$,
indicating the observed F at the anchor exceeds every F obtained at
a random position. This is consistent with the simulated data, which
was constructed with the group difference concentrated at the anchor.

```{r plot-null}
plot_null_distribution(sig)
```

The vertical line at the observed F sits well to the right of the
entire null distribution.

# Alternatives worth knowing about

The methods above are one reasonable choice among several. Depending
on your data and question, one of these might suit you better:

- **Generalized additive models with random effects.** If
  log-linearity is a poor fit and you want to model curvature
  explicitly, `mgcv::gamm` with a smooth term in distance and a
  random slope per sample is a natural extension. The output is
  harder to interpret than an LMM slope but more flexible in shape.
- **Nonlinear mixed models.** If you have a specific mechanistic
  decay function in mind (say, exponential with a plateau), fit that
  form directly with `nlme::nlme`.
- **Genome-wide coverage differential analysis.** If your question
  is really about differences at many positions across the genome
  rather than at a single anchor, tools designed for windowed
  differential coverage analysis
  (`r BiocStyle::Biocpkg("csaw")`,
  `r BiocStyle::Biocpkg("DESeq2")` on binned counts) are more
  appropriate than repeated single-anchor tests.
- **Simple two-stage tests when data is large.** When you have many
  samples per group (say, n > 30), the small-sample corrections
  matter less and a two-stage procedure with permutation-based
  Wilcoxon contrasts may be adequate.

# References

- Bates D, Machler M, Bolker B, Walker S (2015). Fitting linear
  mixed-effects models using lme4. *Journal of Statistical Software*
  67(1):1-48. \doi{10.18637/jss.v067.i01}
- Kenward MG, Roger JH (1997). Small sample inference for fixed
  effects from restricted maximum likelihood. *Biometrics*
  53(3):983-997. \doi{10.2307/2533558}
- Kuznetsova A, Brockhoff PB, Christensen RHB (2017). lmerTest
  package: Tests in linear mixed effects models. *Journal of
  Statistical Software* 82(13):1-26. \doi{10.18637/jss.v082.i13}
- Lenth RV (2024). emmeans: Estimated Marginal Means, aka
  Least-Squares Means. R package.
  <https://CRAN.R-project.org/package=emmeans>
- Phipson B, Smyth GK (2010). Permutation p-values should never be
  zero: calculating exact p-values when permutations are randomly
  drawn. *Statistical Applications in Genetics and Molecular Biology*
  9(1), Article 39. \doi{10.2202/1544-6115.1585}

# Session info

```{r session-info}
sessionInfo()
```
